---
title: "Bayes_approaches"
author: "Wendel Raymond"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r}
library(tidyverse)
library(HDInterval)
library(BayesFactor)
```


## Bayesian appraoches to SAV calculations
Following Hobbs and Hooten texts and code. Bayes Theorem Lab from Bayesian Modeling for Ecologists

You are interested in estimating the posterior distribution for the mean number of individuals of an invasive plant species per m2 in a disturbed grassland (or we can think eelgrass shoots). We will call that mean θ (mean shoot density). You have prior information telling you that the average number of these plants per m2 is 10.2 with a standard deviation of the mean = 0.5. You have a set of fifty observations in hand obtained by sweaty (underwater) labor in the field. Execute the following steps.

### 1 
Simulate 50 data points form a Poisson with mean 7.5. We are calling this our example data. 
```{r}
set.seed(10)

y <- rpois(50, 7.5)
```

### 2 - 3 
Plot
```{r}
hist(y, freq = FALSE, breaks = 10)

mu.prior <- 10.2
sigma.prior <- 0.5
```

```{r}
discrete_hist <- function(y) {
  
  z <- data.frame(y) %>% count(y) %>% mutate(prop = prop.table(n))
  plot(z$y, z$prop, type = "h", ylab = "Probability", xlab = "y", main = "Improved histogram of data", 
  frame = T, xaxt = "n", lwd = 3, col = "blue")
  x <- seq(min(z$y), max(z$y), 1)
  axis(side = 1, at = x, labels = x)

}

discrete_hist(y)
```


### 4
Set up a vector containing a sequence of values for θ, the mean number of invasive (eelgrass shoots) plants.

This where I get a little iffy in that the upper bound here seems to be set arbitrarily, but lets go with it for now.
```{r}
step <- 0.01
theta <- seq(0, 20, step)
```

### 5 - 6
Write the mathematical expression for a gamma prior on θ
```{r}
prior <- function(theta, mu = mu.prior, sigma = sigma.prior) dgamma(theta, mu^2 / sigma^2, mu / sigma^2)

plot(theta, prior(theta = theta, mu.prior, sigma.prior))
```

### 8 - 9
What is the mathematical expression for the likelihood [y∣θ] assuming that the data are conditionally independent?
```{r}
like <- function(theta, y){
  L = rep(0, length(theta))
  for(i in 1:length(theta)) L[i] = prod(dpois(y, theta[i], log = FALSE))
  return(L)
}
```

```{r}
plot(theta, like(theta, y = y), type = "l", xlim = c(0, 20), main = "Likelihood", xlab = expression(theta), 
  ylab = expression(paste("[y|", theta, "]")))
```

### 10
What is the mathematical expression for the joint distribution [θ,y]?
```{r}
joint <- function(theta) like(theta, y = y) * prior(theta)

plot(theta, joint(theta), type = "l",  main = "Joint", xlim = c(0, 20), xlab = expression(theta),
  ylab = expression(paste("[y|", theta, "] x [", theta, "]")))
```

### 11
What is the mathematical expression for the marginal probability of the data [y]?
```{r}
Py <- sum(like(theta, y = y) * prior(theta) * step)
```

### 12
What is the mathematical expression for the posterior distribution [θ∣y]? 
```{r}
p.theta <- joint(theta) / Py

plot(theta, p.theta, typ = "l", xlim = c(0, 20), main = "Posterior", xlab = expression(theta), 
  ylab = expression(paste("[ ", theta, " | y]")))
```

```{r}
par(mfrow = (c(2, 3)))
plot(theta, prior(theta), type = "l", ylab = expression(paste("[", theta, "]")), xlab = expression(theta),
  main = "Prior", xlim = c(0, 20))

hist(y, freq = FALSE, breaks = 10, main = "Histogram of data")
discrete_hist(y = y)

plot(theta, like(theta, y = y), type = "l", main = "Likelihood", xlim = c(0, 20), xlab = expression(theta),
  ylab = expression(paste("[y|", theta, "])")))
plot(theta, joint(theta), type = "l", main = "Joint", xlim = c(0, 20), xlab = expression(theta),
  ylab = expression(paste("[y | ", theta, "]) x [", theta, "]")))
plot(theta, p.theta, type = "l", xlim = c(0, 20), main = "Posterior", xlab = expression(theta),
  ylab = expression(paste("[ ", theta, " | y]")))
```

### 14
```{r}
(c <- max(p.theta) / max(like(theta, y)))

like.scaled <- c * like(theta, y)

par(mfrow=c(1, 1))
plot(theta, like.scaled, type = "l", col = "red", xlim = c(0, 15), xlab = expression(theta),
  main = "Scaled Overlay", ylab = "Probability density")
lines(theta, p.theta, type = "l")
lines(theta, prior(theta), col = "blue")
legend(0, 1, c("Scaled likelihood","Posterior", "Prior"), lty = rep("solid",3), col = c("red", "black", "blue"))
```

### More realistic eelgrass example

#### Data
```{r}
## Prior data from preliminary sampling ##
mu.prior <- 20
sigma.prior <- 10

## simulated new sampled data (e.g. 1 year post construction) ##
y <- rpois(50, 10)
```

#### Parameters and probabilities
```{r}
## Theta: mean number of shoots ##
step <- 0.1
theta <- seq(0, 40, step)

## Prior distribution ##
prior <- function(theta, mu = mu.prior, sigma = sigma.prior) dgamma(theta, mu^2 / sigma^2, mu / sigma^2)
plot(theta, prior(theta = theta, mu.prior, sigma.prior))

## Liklihood: probability of theta given the data ##
like <- function(theta, y){
  L = rep(0, length(theta))
  for(i in 1:length(theta)) L[i] = prod(dpois(y, theta[i], log = FALSE))
  return(L)
}

plot(theta, like(theta, y = y), type = "l", xlim = c(0, 20), main = "Likelihood", xlab = expression(theta), 
  ylab = expression(paste("[y|", theta, "]")))

## Joint ##
joint <- function(theta) like(theta, y = y) * prior(theta)
plot(theta, joint(theta), type = "l",  main = "Joint", xlim = c(0, 20), xlab = expression(theta),
  ylab = expression(paste("[y|", theta, "] x [", theta, "]")))

## Marginal probability ##
Py <- sum(like(theta, y = y) * prior(theta) * step, na.rm = TRUE)

## Posterior ##
p.theta <- joint(theta) / Py
plot(theta, p.theta, typ = "l", xlim = c(0, 20), main = "Posterior", xlab = expression(theta), 
  ylab = expression(paste("[ ", theta, " | y]")))
```

#### Plots
```{r}
## Grid ##
par(mfrow = (c(2, 3)))

plot(theta, prior(theta), type = "l", ylab = expression(paste("[", theta, "]")), xlab = expression(theta),
  main = "Prior")

hist(y, freq = FALSE, breaks = 10, main = "Histogram of data")

discrete_hist(y = y)

plot(theta, like(theta, y = y), type = "l", main = "Likelihood", xlab = expression(theta), ylab = expression(paste("[y|", theta, "])")))

plot(theta, joint(theta), type = "l", main = "Joint", xlab = expression(theta), ylab = expression(paste("[y | ", theta, "]) x [", theta, "]")))

plot(theta, p.theta, type = "l", main = "Posterior", xlab = expression(theta), ylab = expression(paste("[ ", theta, " | y]")))

## Single overlay ##
(c <- max(p.theta, na.rm = TRUE) / max(like(theta, y), na.rm = TRUE))

like.scaled <- c * like(theta, y)

par(mfrow=c(1, 1))
plot(theta, like.scaled, type = "l", col = "red", xlim = c(0, 40), xlab = expression(theta),
  main = "Scaled Overlay", ylab = "Probability density")
lines(theta, p.theta, type = "l")
lines(theta, prior(theta), col = "blue")
legend(2, 0.6, c("Scaled likelihood","Posterior", "Prior"), lty = rep("solid",3), col = c("red", "black", "blue"))
```

### With rmBayes

#### Simulate data
```{r}
dat <- data.frame(time = c(rep("init", 30), rep("yr1", 30)),
                  shoots = c(rpois(30, 20), rpois(30, 10)))

mu_init <- mean(dat$shoots[dat$time=="init"])
sigma_inti <- sd(dat$shoots[dat$time=="init"])
mu_yr1 <- mean(dat$shoots[dat$time=="yr1"])
sigma_yr1 <- sd(dat$shoots[dat$time=="yr1"])
```

#### Using BayesFactor
```{r}
bf_rmanova <- anovaBF(shoots ~ time, whichRandom = "time", data = dat)
bf_rmanova
```


