---
title: "SAV_Sample_Size"
author: "Wendel Raymond"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)

theme_set(theme_classic())
```


## Submerged Aquatic Vegitation Sample Size Calculations
Calcualte sample size needed to detect a 20% difference in SAV counts between two groups. Explore with various power values.

### Dummy data
Using Wendel's dissertation data from 21 sites in SE Alaska. N=8 0.25 m-sq quadrats. These data are helpful, but its important to consider that real world data will probably be more zero and low number rich. 

```{r}
eg.dat <- read.csv("Puyallup_MSA_Eelgrass.csv", stringsAsFactors = FALSE, header = TRUE)
```

### Summarize data
```{r}
eg.sum <- eg.dat %>% 
  group_by(Tjoin) %>% 
  summarise(n_samples = n(),
            n_EG_pres = sum(Eg_PA),
            mean_EG_dens = mean(Eg_dens, na.rm = TRUE))

eg.sum$n_EG_dens_orig <- c(4, 0, 3, 2, 0, 1, 2, 2, 1, 0)
eg.sum$n_to_sim <- eg.sum$n_EG_pres - eg.sum$n_EG_dens_orig
```

### Simulate data
Use mean values from each transect
```{r}
eg.sim <- data.frame(Tjoin = rep(eg.sum$Tjoin, eg.sum$n_samples))
eg.sim$rep <- c(1:eg.sum$n_samples[1],
                1:eg.sum$n_samples[2],
                1:eg.sum$n_samples[3],
                1:eg.sum$n_samples[4],
                1:eg.sum$n_samples[5],
                1:eg.sum$n_samples[6],
                1:eg.sum$n_samples[7],
                1:eg.sum$n_samples[8],
                1:eg.sum$n_samples[9],
                1:eg.sum$n_samples[10])
eg.sim$EG_PA <- eg.dat$Eg_PA

t1 <- rpois(n = eg.sum$n_to_sim[1], lambda = eg.sum$mean_EG_dens[1])
t2 <- rpois(n = eg.sum$n_to_sim[3], lambda = eg.sum$mean_EG_dens[3])
t3 <- rpois(n = eg.sum$n_to_sim[4], lambda = eg.sum$mean_EG_dens[4])
t5 <- rpois(n = eg.sum$n_to_sim[6], lambda = eg.sum$mean_EG_dens[6])
t6 <- rpois(n = eg.sum$n_to_sim[7], lambda = eg.sum$mean_EG_dens[7])
t7 <- rpois(n = eg.sum$n_to_sim[8], lambda = mean(eg.sum$mean_EG_dens, na.rm = TRUE))
t8 <- rpois(n = eg.sum$n_to_sim[9], lambda = mean(eg.sum$mean_EG_dens, na.rm = TRUE))

eg.sim$EG_dens <- NA
eg.sim[1:23, 4] <- c(37.33, 37.33, 37.33, 0, t1, 0, 0, 0, 0, 0 ,0 ,0, 0, 0, 0)
eg.sim[24:41, 4] <- rep(0, 18)
eg.sim[42:63, 4] <- c(53.33, 0, 26.67, t2, rep(0, 11))
eg.sim[64:85, 4] <- c(0, 16, t3, rep(0, 15))
eg.sim[86:102, 4] <- rep(0, 17)
eg.sim[103:123, 4] <- c(10.67, t5, rep(0, 17))
eg.sim[124:145, 4] <- c(42.67, 0, t6, rep(0, 15))
eg.sim[146:165, 4] <- c(t7, rep(0, 15))
eg.sim[166:184, 4] <- c(t8, rep(0, 16))
eg.sim[185:201, 4] <- rep(0, 17)
```

### Calculate sample size
Calculate sample size needed to detect a 20% change between two groups. alpha = 0.1, beta = 0.9
```{r}
## All samples ##
mean(eg.sim$EG_dens)
sd(eg.sim$EG_dens)

(2*(qnorm(0.9) + qnorm(0.8))^2) * ((sd(eg.sim$EG_dens) / (mean(eg.sim$EG_dens) * 0.2))^2)

## Ilai's transformation ##
(abs((8/0.04) * qnorm(0.1))) * (sd(eg.sim$EG_dens) / mean(eg.sim$EG_dens)^2)

## Eelgrass present ##
mean(eg.sim$EG_dens[eg.sim$EG_PA == 1])
sd(eg.sim$EG_dens[eg.sim$EG_PA == 1])

(2*(qnorm(0.9) + qnorm(0.8))^2) * ((sd(eg.sim$EG_dens[eg.sim$EG_PA == 1]) / (mean(eg.sim$EG_dens[eg.sim$EG_PA == 1]) * 0.2))^2)

## Ilai's transformation ##
(abs((8/0.04) * qnorm(0.1))) * (sd(eg.sim$EG_dens[eg.sim$EG_PA == 1]) / mean(eg.sim$EG_dens[eg.sim$EG_PA == 1])^2)
```

