---
title: "Worked_example"
author: "Wendel Raymond"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Worked example
Example data and analysis for an eelgrass analysis of a HPA permitted project. Write up in Appendix A of Eelgrass and macroalgae survey guidelines.

```{r}
library(tidyverse)
theme_set(theme_classic())
set.seed(9)
```

### Data
```{r}
dat <- read.csv("example_eelgrass_data.csv", header = TRUE, stringsAsFactors = FALSE)

dat$time <- factor(dat$time, levels = c("pre", "one", "three", "five"))
dat$site <- factor(dat$site, levels = c("proj", "ref"))

dat.sum <- dat %>% 
  group_by(site, time) %>% 
  summarize(mean = mean(density),
            sd = sd(density))

pal <- c("#6A8026", "#3542A1")

manu.dat <- data.frame(
  quad = seq(1:50),
  eg = c(rpois(n = 18, lambda = 30), rep(0, 32))
)
```

690 total survey feet, 338 have eelgrass so 48.9% of quads should have eelgrass for a simulation or 389 quads

```{r}
dat.pre <- data.frame(
  quad = seq(1:796),
  density = c(rpois(n = 389, lambda = 24), rep(0, 407)),
  time = "pre",
  site = "proj"
)

dat.post1 <- data.frame(
  quad = seq(1:796),
  density = c(rpois(n = 389, lambda = 23), rep(0, 407)),
  time = "post_1",
  site = "proj"
)

dat.post3 <- data.frame(
  quad = seq(1:796),
  density = c(rpois(n = 250, lambda = 28), rep(0, 546)),
  time = "post_3",
  site = "proj"
)

dat.post5 <- data.frame(
  quad = seq(1:796),
  density = c(rpois(n = 250, lambda = 26), rep(0, 546)),
  time = "post_5",
  site = "proj"
)

dat.pre.ref <- data.frame(
  quad = seq(1:796),
  density = c(rpois(n = 389, lambda = 24), rep(0, 407)),
  time = "pre",
  site = "ref"
)

dat.post1.ref <- data.frame(
  quad = seq(1:796),
  density = c(rpois(n = 389, lambda = 23), rep(0, 407)),
  time = "post_1",
  site = "ref"
)

dat.post3.ref <- data.frame(
  quad = seq(1:796),
  density = c(rpois(n = 290, lambda = 28), rep(0, 506)),
  time = "post_3",
  site = "ref"
)

dat.post5.ref <- data.frame(
  quad = seq(1:796),
  density = c(rpois(n = 340, lambda = 21), rep(0, 456)),
  time = "post_5",
  site = "ref"
)

dat <- rbind.data.frame(dat.pre, dat.post1, dat.post3, dat.post5, dat.pre.ref, dat.post1.ref, dat.post3.ref, dat.post5.ref)

dat$density <- ifelse(dat$density < 0, 0, dat$density)

dat$time <- factor(dat$time, levels = c("pre", "post_1", "post_3", "post_5"))

dat.sum <- dat %>% 
  group_by(site, time) %>% 
  summarize(mean = mean(density),
            sd = sd(density))
```


### N-star
```{r}
# Known parameters
mean_pre <- round(mean(manu.dat$eg), 0)
alpha <- 0.1      # Significance level
beta <- 0.9       # Desired power
sd_pre <- round(sd(manu.dat$eg), 0)
delta <- (0.2 * mean_pre)  # Minimum detectable effect size (20% change)

# Calculate critical values
z_alpha_over_2 <- qnorm(1 - alpha/2)
z_beta <- abs(qnorm(1 - beta))

# Calculate sample size per group
sample_size <- ((2 * (z_alpha_over_2 + z_beta)^2 * sd_pre^2) / delta^2)
round(sample_size, 0)

200*qnorm(0.1)^2 / 5

2 *(qnorm(0.1) + qnorm(0.9))^2 * (10 / (10 - 8)^2)

## Whitehead
SSizWhitehead <- function(alpha, power, lambda_1, lambda_2, ratio) # estimate sample size
{
  tr = abs(-log(lambda_1/lambda_2))      
  lb = (ratio * lambda_1 + lambda_2)/(ratio + 1)
  za = qnorm(alpha)        # 1 tail
  zb = qnorm(1 - power)
  v = ((za + zb) / tr)^2
  return (round(v / ratio / lb * (ratio + 1) * (ratio + 1))) # total sample size (2 groups)
}

SSizWhitehead(0.1, 0.8, 11, (11 * 0.8), 1)
```

### Monitoring and mitigation

### N* transect allocation
```{r}
tran.dat <- data.frame(
  transect = as.character(1:7),
  length = c(90, rep(100, 6))
)

tran.dat$prop_tot_len <- tran.dat$length / sum(tran.dat$length)

tran.dat$tran_N <- round(tran.dat$prop_tot_len * round(sample_size, 0), 0)
```

#### Option 1: Upfront mitigation

```{r}
mean_pre <- dat.sum[[1, 3]]
sd_pre <- dat.sum[[1, 4]]

surv_area <- ((90 * 10) + (6 * 100 * 10)) / 10.764
eg_area <- ( (46 * 10) +
             (38 * 10) +
             (34 * 10) +
             (37 * 10) +
             (60 * 10) +
             (64 * 10) +
             (59 * 10)) / 10.764

tot_pre_shoots <- eg_area * mean_pre
round(tot_pre_shoots, 0)
```

#### Option 2: No upfront mitigation

```{r}
png(filename = "../Figures/worked_example_proj.png", width = 5, height = 3.5, units = "in", res = 400)
ggplot(dat.sum %>% 
         filter(site == "proj")) +
  geom_col(aes(x = time, y = mean, fill = site)) +
  geom_errorbar(aes(x = time, ymin = mean - sd, ymax = mean + sd, group = site), width = 0) +
  scale_fill_manual(values = pal, name = "Site", breaks = c("proj", "ref"), labels = c("Project", "Reference")) +
  geom_hline(aes(yintercept = 11.8), color = pal[1], linetype = "dashed") +
  xlab("Time") +
  ylab("Shoot density (m-sq) +/- SD") +
  theme(text = element_text(size = 12)) +
  theme(legend.position = "none")
dev.off()

ggplot(dat %>% 
         filter(site == "proj")) +
  geom_histogram(aes(x = density)) +
  facet_wrap(~ time)
```

##### Analyses
```{r}
## Year 1 ##
dat.1 <- dat %>% 
  filter(site == "proj") %>% 
  filter(time %in% c("pre", "post_1"))

mod.1 <- glm(density ~ time, data = dat.1, family = quasipoisson(link = "log"))
summary(mod.1)

## Year 3 ##
dat.2 <- dat %>% 
  filter(site == "proj") %>% 
  filter(time %in% c("pre", "post_3"))

mod.2 <- glm(density ~ time, data = dat.2, family = quasipoisson(link = "log"))
summary(mod.2)

## Year 5 ##
### Compare to pre ##
dat.3 <- dat %>% 
  filter(site == "proj") %>% 
  filter(time %in% c("pre", "post_5"))

mod.3 <- glm(density ~ time, data = dat.3, family = quasipoisson(link = "log"))
summary(mod.3)

### Compare to year 3 ##
dat.4 <- dat %>% 
  filter(site == "proj") %>% 
  filter(time %in% c("post_3", "post_5"))

mod.4 <- glm(density ~ time, data = dat.4, family = quasipoisson(link = "log"))
summary(mod.4)
```

##### Mitigation

```{r}
## Year 3 ##
mean_3yr <- exp(2.465 - 0.288)

eg_shoots_3yr <- eg_area * mean_3yr

mit_shoots_3yr <- tot_pre_shoots - eg_shoots_3yr
round(mit_shoots_3yr)
```

#### Option 3: No upfront mitgation with reference.

```{r}
png(filename = "../Figures/worked_example_proj_ref.png", width = 5.5, height = 3.5, units = "in", res = 400)
ggplot(dat.sum) +
  geom_col(aes(x = time, y = mean, group = site, fill = site), position = position_dodge(1)) +
  geom_errorbar(aes(x = time, ymin = mean - sd, ymax = mean + sd, group = site), position = position_dodge(1), width = 0) +
  scale_fill_manual(values = pal, name = "Site", breaks = c("proj", "ref"), labels = c("Project", "Reference")) +
  geom_hline(aes(yintercept = 11.8), color = pal[1], linetype = "dashed") +
  geom_hline(aes(yintercept = 11.6), color = pal[2], linetype = "dashed") +
  xlab("Time") +
  ylab("Shoot density (m-sq) +/- SD") +
  theme(text = element_text(size = 12))
dev.off()
```

##### Analyses

```{r}
## Project to reference ##
dat.5 <- dat %>% 
  filter(site %in% c("proj", "ref")) %>% 
  filter(time %in% c("pre"))

mod.5 <- glm(density ~ site, data = dat.5, family = quasipoisson(link = "log"))
summary(mod.5)

## Year 1 ##
dat.6 <- dat %>% 
  filter(time %in% c("pre", "post_1"))

mod.6 <- glm(density ~ time/site, data = dat.6, family = quasipoisson(link = "log"))
summary(mod.6)

## Year 3 ##
dat.7 <- dat %>% 
  filter(time %in% c("pre", "post_3"))

mod.7 <- glm(density ~ time/site, data = dat.7, family = quasipoisson(link = "log"))
summary(mod.7)

## Year 5 ##
dat.8 <- dat %>% 
  filter(time %in% c("pre", "post_5"))

mod.8 <- glm(density ~ time/site, data = dat.8, family = quasipoisson(link = "log"))
summary(mod.8)

### Compare to year 3 ##
dat.9 <- dat %>% 
  filter(time %in% c("post_3", "post_5"))

mod.9 <- glm(density ~ time/site, data = dat.9, family = quasipoisson(link = "log"))
summary(mod.9)
```

##### Mitigation
```{r}
## Year 3 ##
mean_3yr_ref <- exp(mod.7$coefficients[[1]] - mod.7$coefficients[[4]])

tot_shoots_3yr_ref <- eg_area * mean_3yr_ref

mit_shoots_3yr_ref <- tot_pre_shoots - tot_shoots_3yr_ref
round(mit_shoots_3yr_ref)
```

### Other examples

#### Ex 1

edit data 
```{r}
dat.adj.1 <- dat %>% 
  filter(time %in% c("pre", "post_1")) 

dat.adj.1[797:1107, 2] <- dat.adj.1[797:1107, 2] * 0.75
dat.adj.1[2389:2698, 2] <- dat.adj.1[2389:2698, 2] * 0.75

dat.adj.1.sum <- dat.adj.1 %>% 
  group_by(site, time) %>% 
  summarise(mean = mean(density),
            sd = sd(density))
```

```{r}
mod.10 <- glm(density ~ time/site, data = dat.adj.1, family = quasipoisson(link = "log"))
summary(mod.10)
```

```{r}
ref_adj_shoots <- exp(mod.10$coefficients[[1]] + (mod.10$coefficients[[2]] + 0.200))

shoots_ref_adj <-  -(mean_pre - ref_adj_shoots * 314)

mit_shoots_ref_adj <- tot_pre_shoots - shoots_ref_adj
round(mit_shoots_ref_adj)

## with no ref ##
tot_pre_shoots - (tot_area * dat.adj.1.sum[[2, 3]])
```

